---
title: "Assignment 5: Data Visualization"
author: "Jalal Bayar"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Loading packages
library(tidyverse); library(lubridate); library(here); library(cowplot)
getwd()
Processed_data <- ("./Data/Processed_KEY")

#read in data files

Peter.Paul.chem.nutrient <- read.csv(
  here(Processed_data, "NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
       stringsAsFactors = TRUE)

Litter <- read.csv(
  here(Processed_data, "NEON_NIWO_Litter_mass_trap_Processed.csv"),
       stringsAsFactors = TRUE)

#2 command for date as Date object
Peter.Paul.chem.nutrient$sampledate <- ymd(Peter.Paul.chem.nutrient$sampledate)

Litter$collectDate <- ymd(Litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme} 
#
library(ggridges)

#Enhanced color ramps
library(viridis)
library(RColorBrewer)
library(colormap)

#Add pre-set themes <--NEW
library(ggthemes)
#3

mytheme <- theme_classic() + 
  theme(
    line = element_line(
      color='red',
      linewidth = 1
    ),
    legend.background = element_rect(
      color='grey',
      fill = 'white'
    ),
    legend.title = element_text(
      color='black'
    ),
    legend.position = "top"
  )


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
#4 Function for Scatter plot and use 'ylim' to limit y-axis. Also, some other 
#commands such as xlab and ylab to give a name to x and y-axis. Geom-smooth to 
#draw a line(s).

Peter.Paul.TotalP.po4 <- ggplot(Peter.Paul.chem.nutrient,
                                aes(x = tp_ug, y = po4, color = lakename)) +
  geom_point() +
  ylim(0, 45) +
  labs(color = "Lake") +
  geom_smooth(method = lm) +
  xlab("Totol P(ug)") +
  ylab("Phosphate") +
  mytheme
print(Peter.Paul.TotalP.po4)


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 Boxplot 
# I draw two different commands for each variables to see which function is more visually more effective. 

Peter.Paul.temp <- ggplot(Peter.Paul.chem.nutrient) +
  geom_boxplot(
    aes(x=month, 
        y = temperature_C, 
        color = lakename)) +
  mytheme +
  labs(color = "Lake") +
  xlab(expression()) 
  print(Peter.Paul.temp)

Peter.Paul.Total.P <- ggplot(Peter.Paul.chem.nutrient) +
  geom_boxplot(
    aes(x=month, 
        y = tp_ug, 
        color = lakename)) +
    mytheme +
  theme(legend.position = "none") +
  xlab(expression())
  print(Peter.Paul.Total.P)

Peter.Paul.Total.N <- ggplot(Peter.Paul.chem.nutrient) +
  geom_boxplot(
    aes(x=month, 
        y = tn_ug, 
        color = lakename)) +
  mytheme +
  theme(legend.position = "none")
  print(Peter.Paul.Total.N)


Combined.Peter.Paul.Temp.P.N <- plot_grid(Peter.Paul.temp + ylab("Temp_C"),
                                Peter.Paul.Total.P, 
                                Peter.Paul.Total.N, nrow = 3)
print(Combined.Peter.Paul.Temp.P.N)

# A duplicate graph to see individual months distribution of the required varibles.

Peter.Paul.Temp_duplicate <- ggplot(Peter.Paul.chem.nutrient) +
  geom_boxplot(
    aes(x=factor(month,levels=1:12,labels=month.abb), 
        y = temperature_C, 
        color = lakename)) +
  mytheme +
  labs(color = "Lake") +
  xlab(expression())
  print(Peter.Paul.Temp_duplicate)  
  
Peter.Paul.Total.P_duplicate <- ggplot(Peter.Paul.chem.nutrient) +
  geom_boxplot(
    aes(x=factor(month,levels=1:12,labels=month.abb), 
        y = tp_ug, 
        color = lakename)) +
    mytheme +
  theme(legend.position = "none") +
  xlab(expression()) 
  print(Peter.Paul.Total.P_duplicate)
  
Peter.Paul.Total.N_duplicate <- ggplot(Peter.Paul.chem.nutrient) +
  geom_boxplot(
    aes(x=factor(month,levels=1:12,labels=month.abb), 
        y = tn_ug, 
        color = lakename)) +
  mytheme +
  theme(legend.position = "none") +
  xlab("Month")
  print(
Peter.Paul.Total.N_duplicate)

Combined.Peter.Paul.Temp.P.N_duplicate <- plot_grid(Peter.Paul.Temp_duplicate + ylab("Temp_C"),
                             Peter.Paul.Total.P_duplicate, 
                             Peter.Paul.Total.N_duplicate, nrow = 3)
print(Combined.Peter.Paul.Temp.P.N_duplicate)


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Overall, the concentration of TP is ranging from 0 to 150 while the
concentration of TN ranging from 0 to 3000 for both the Paul and Peter lake. The 
temp_C ranged from 0 to around 25 degree C. The data distibution is more condensed 
and abundant at the concentration 0 to 30 ug for TP, and 0 to 400 ug for TN across 
three month period. The temp_C were observed over 5 month period. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
Litter.needles <- ggplot(subset(Litter, 
  functionalGroup == "Needles"),
  aes(x = collectDate, y = dryMass, color = nlcdClass)) +
    geom_boxplot() +
  labs(color = "Sample Type") +
  mytheme
print(Litter.needles)

#7 In the following command, I removed the legend because we can see different facets with legend's name.

Litter.needles.facet <- ggplot(subset(Litter, 
  functionalGroup == "Needles"),
  aes(x = collectDate, y = dryMass, color = nlcdClass)) +
    geom_boxplot(size = 1) +
  facet_wrap(facets = vars(nlcdClass), nrow=1,ncol=3) +
  ylim(0, 40) +
  xlab("Year") +
  mytheme +
  theme(legend.position = "none")
print(Litter.needles.facet)


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 is more effective because we can see clearly the three sampling sites (facets) with the distribution in each year. However, in case of plot 6, we see the distribution across three years as a whole not separately. Plot 6 cannot gives us complete information.