---
title: "Assignment 3: Data Exploration"
author: "Jalal Bayar"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
library(tidyverse) # Loading tidyverse, lubridate, and here libraries.
library(lubridate)
library(here)
getwd() # checking working directry
Neonics <- read.csv(
  file = here("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"),
  stringsAsFactors = TRUE) # Importing Neonics data

Litter <- read.csv(
  file = here("./Data/Raw/NIWO_Litter/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE
) # Importing litter data

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: I think, through this database, we can visualize data from multiple studies that have been used different exposures, animals, durations, and endpoints etc. It is helpful to make a correlation between different studies that are there any commonalities between different studies with achieving different objectives and answering various research questions? Different insecticides could have adverse outcomes for multiple species including humans.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter and woody debris play an important role in nutrient cycling and carbon sequestration. Also, the decomposition rate of litter and woody can vary accross different climates such as temperature and humdidity play a crucial role. We can predict models how litter and debris fall can contribute to the ecosystem as well as how it could be affected by climate change.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: NEON collects data accross the U.S. to better understand the ecosystem. Both terrestrial and aquatic field sites are located where litter and debris collection represent terrestrial ecosystem. The network focuses on litter and woody debris that regulates the nutrient cycling, carbon sequestration, and habitat creation. NEON collects data across the continent based on a standarized method to better understand the composition and decomposition of litter and debris. Sampling of the litter and debris occurs in tower plots. The location of tower plots are selected randomly within the 90% flux footprint of the primary and secondary airsheds. The three key factors that affect the sampling are as follows
 1. Plots edges must be separated by a distance 150% of one edge of the plot.
 2. Plot centers must be greater than 50m from large paved roads.
 3. Plot edges must be 10m from two-tract dirt roads.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
# dim command is used to see the total observations and variables in a data set.
dim(Neonics) # 4623 30

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
summary(Neonics$Effect)
sort(summary(Neonics$Effect))
```

> Answer: The highest effect that has been found is polulatiion and mortality which might means that neonicotinides are extremely hazardous to health. Also, other effects such as  behavior, reproduction, and development of various insects have been studied. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
summary(Neonics$Species.Common.Name, maxsum = 7) # Honey bee, Parasite Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, Italian Honey Bee.

```

> Answer: The above mentioned insects play an important role in pollination. These are studied to see how neonicotinoids have adverse effect on the population, growth, reproduction or on other factors.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
class(Neonics$Conc.1..Author.) # Factor

```

> Answer: Because it is a group of varibles with in the concentration column. Also, some values have (/) signs and are NR.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year))
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location))
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: There are four test location, field artificial, field natural, field underminable, and Lab. The most common location is Lab followed by field natural. Yes, there are variation in test location over time.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
summary(Neonics$Endpoint, maxsum = 7) # 28 end points.
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint)) # command for bar garph
 
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint)) +   
           theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) # command for aligning and rotating x-axis labels with bar graph command.

```

> Answer: The most common observations Lowest observed effect level (LOEL) and No observed effect level (NOEL). LOEL is the lowest concentration of the effect where results are statistically significant, however, NOEL is the highest concentation where there is no significant correlation in the concentation being tested.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)

Litter$collectDate <- ymd(Litter$collectDate) # convert date as factor to Date as object.

class(Litter$collectDate)

unique(Litter$collectDate) # "2018-08-02" "2018-08-30"

```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
summary(Litter$plotID)

ggplot(Litter) +
  geom_bar(aes(x = plotID)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
unique(Litter$plotID) 
```

> Answer: The summary function gives me all the observation in each of the sample plot (total 12), however, the unique function exclude the sample plot that has the same sample plot numbers.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter) +
  geom_bar(aes(x = functionalGroup))


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass))
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) # command for boxplot

ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass))

ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) # command for violen plot

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The box plot shows the min, max, interquartile range, and potential outliers, however, violen plot shows the distribution of different variables in addition to the characteristics of box plot. In this case, box plot is more visually representative, we can clearly identify min, median ann IQR of all the variables, but violen plot is not giving us the more representive visualization becuase it incorporate more indepth distribution of data in plots and the distribution of data is not applicable to violen plot.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles would have the highest biomass compared to other variables. 
