---
title: "Assignment 8: Time Series Analysis"
author: "Jalal Bayar"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r}
library(tidyverse); library(lubridate); library(zoo); library(trend)
library(here)
getwd()

mytheme <- theme_classic(
  base_size = 14) +
  theme(axis.text = element_text(color = "black"),
legend.position = "top")
theme_set(mytheme)


```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1
EPA_2010 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv"),
                     stringsAsFactors = TRUE)
EPA_2011 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv"),
                     stringsAsFactors = TRUE)
EPA_2012 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv"),
                     stringsAsFactors = TRUE)
EPA_2013 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv"),
                     stringsAsFactors = TRUE)
EPA_2014 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv"),
                     stringsAsFactors = TRUE)
EPA_2015 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv"),
                     stringsAsFactors = TRUE)
EPA_2016 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv"),
                     stringsAsFactors = TRUE)
EPA_2017 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv"),
                     stringsAsFactors = TRUE)
EPA_2018 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv"),
                     stringsAsFactors = TRUE)
EPA_2019 <- read.csv(
  file = here("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv"),
                     stringsAsFactors = TRUE)

GaringerOzone <- rbind(EPA_2010, EPA_2011, EPA_2012, EPA_2013, EPA_2014,
                       EPA_2015, EPA_2016, EPA_2017, EPA_2018, EPA_2019)

```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3 # using lubridate function
GaringerOzone$Date <- mdy(GaringerOzone$Date)

# 4 # using select function to wrangle data
GaringerOzone <- select(GaringerOzone, "Date",
                        "Daily.Max.8.hour.Ozone.Concentration",
                        "DAILY_AQI_VALUE")

# 5 creating a new sequence of dates for 10 years
date_sequence <- seq(from = as.Date("2010-01-01"), to = as.Date("2019-12-31"), by = "day")

Days <- as.data.frame(date_sequence)

colnames(Days) <- c("Date")

# 6 comibining two dataframes with left_join command
combined_data <- left_join(Days, GaringerOzone)

GaringerOzone <- combined_data

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7

ggplot(GaringerOzone, aes(x = Date, y = Daily.Max.8.hour.Ozone.Concentration)) +
  geom_line() +
  geom_smooth(method = lm) +
  ylab("Ozone (ppm)") +
  xlab("Year") +
  scale_x_date(date_breaks = "1 years", date_labels = "%Y") +
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))

```

>Answer:Yes, the Ozone concentration decreased over time, but also a slight increase after 2019. 

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8
  
GaringerOzone_clean <- GaringerOzone %>% 
  mutate(Ozone = zoo::na.approx(Daily.Max.8.hour.Ozone.Concentration))

```

> Answer: Piecewise constant will assumed the same values earlier or later near to the neighbor, however, ozone concentration change over time. Also, spline can be used for quadratic relationship not linear.


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9
GaringerOzone.monthly <- GaringerOzone_clean %>% 
  mutate(Month = month(Date), Year = year(Date)) %>% 
  group_by(Month, Year) %>% 
  summarise(mean_ozone = mean(Ozone), .groups = 'drop') %>%
  mutate(Date = my(paste0(Month,"-",Year)))

```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10

GaringerOzone.daily.ts <- ts(GaringerOzone_clean$Ozone,
                             frequency = 12,
                             start = c(2010,1)
                             )
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$mean_ozone,
                               frequency = 12,
                               start = c(2010,1))

```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11
Daily_O_decomposed <- stl(GaringerOzone.daily.ts,
                          s.window = "periodic")
plot(Daily_O_decomposed)

Monthly_O_decomposed <- stl(GaringerOzone.monthly.ts,
                            s.window = "periodic")
plot(Monthly_O_decomposed)

```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12
Monthly_trend <- trend::smk.test(GaringerOzone.monthly.ts)
Monthly_trend

```

> Answer: We use Mann-Kendall trend because our data is non-parametric -following not specific distribution- and seasonal.

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}
# 13
ggplot(GaringerOzone.monthly,
       aes(x = Date, y = mean_ozone)) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "Ozone") +
  scale_x_date(date_breaks = "1 years", date_labels = "%Y") +
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))

```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: Overall, from 2010 to 2011, the ozone concentrtion did not change. The mean ozone in 2010 and 2022 is ~0.03. However, the distribution of ozone gradually high in the start of year and mostly at the peak in the middle of the year, and then gradually decrease in the later part of the year (p-value > 0.05).

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15
GaringerOzone.monthly.ts_decomposed <- stl(GaringerOzone.monthly.ts, s.window = "periodic")

Ozone.monthly.components <- as.data.frame(GaringerOzone.monthly.ts_decomposed$time.series[,1:3])

seasonal_component <- GaringerOzone.monthly.ts_decomposed$time.series[, "seasonal"]

deseasonalized_ts <- GaringerOzone.monthly.ts - seasonal_component

plot(GaringerOzone.monthly.ts_decomposed)

#16

library(Kendall)

non.seasonal.trend.monthly.ozone <- MannKendall(deseasonalized_ts)

non.seasonal.trend.monthly.ozone

```

> Answer: Both the results are non significant. The p-value is 0.1712 when we run seasonal Mann-Kendall trend test and p-value is 0.10388 with non-seasonal test. It means seasonality play a role to some extend in distribution of monthly ozone concentration from 2010 to 2019. 
