---
title: "Assignment 4: Data Wrangling (Fall 2024)"
author: "Jalal Bayar"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.
6. Ensure that code in code chunks does not extend off the page in the PDF.


## Set up your session 

1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.

1b. Check your working directory.

1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Add the appropriate code to reveal the dimensions of the four datasets. 

```{r setup, message = FALSE}
#1a Loading libraries
library(tidyverse)
library(lubridate)
library(here)
#1b checking working directory
getwd()
#1c importing files # I commented line 40 to 43 not to include the output in the pdf.
#read.csv(file = here("./Data/Raw/EPAair_O3_NC2018_raw.csv"),stringsAsFactors = TRUE)
#read.csv(file = here("./Data/Raw/EPAair_O3_NC2019_raw.csv"),stringsAsFactors = TRUE)
#read.csv(file = here("./Data/Raw/EPAair_PM25_NC2018_raw.csv"),stringsAsFactors = TRUE)
#read.csv(file = here("./Data/Raw/EPAair_PM25_NC2019_raw.csv"),stringsAsFactors = TRUE)

#2 assigning names
ozone.air.2018 <- read.csv(
  file = here("./Data/Raw/EPAair_O3_NC2018_raw.csv"),
  stringsAsFactors = TRUE
)

ozone.air.2019 <- read.csv(
  file = here("./Data/Raw/EPAair_O3_NC2019_raw.csv"),
  stringsAsFactors = TRUE
)

pm25.air.2018 <- read.csv(
  file = here("./Data/Raw/EPAair_PM25_NC2018_raw.csv"),
  stringsAsFactors = TRUE
)

pm25.air.2019 <- read.csv(
  file = here("./Data/Raw/EPAair_PM25_NC2019_raw.csv"),
  stringsAsFactors = TRUE
)
# checking no of variables and observations in all the data sets
dim(ozone.air.2018) 
dim(ozone.air.2019)
dim(pm25.air.2018)
dim(pm25.air.2019)

```

>All four datasets should have the same number of columns but unique record counts (rows). Do your datasets follow this pattern?


## Wrangle individual datasets to create processed files.

3. Change the Date columns to be date objects.

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r questions 3 thru 4}
#3 conversion of date as factor to date as object

ozone.air.2018$Date <- as.Date(ozone.air.2018$Date, format = "%m/%d/%Y")
ozone.air.2019$Date <- as.Date(ozone.air.2019$Date, format = "%m/%d/%Y")
pm25.air.2018$Date <- as.Date(pm25.air.2018$Date, format = "%m/%d/%Y")
pm25.air.2019$Date <- as.Date(pm25.air.2019$Date, format = "%m/%d/%Y")

#4 Making a subset of a required columns for assignment with select command.
colnames(ozone.air.2018)

ozone.air.2018.subset <- select(ozone.air.2018, "Date", "DAILY_AQI_VALUE", "Site.Name", "AQS_PARAMETER_DESC" , "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")


ozone.air.2019.subset <- select(ozone.air.2019, "Date", "DAILY_AQI_VALUE", "Site.Name", "AQS_PARAMETER_DESC" , "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")


pm25.air.2018.subset <- select(pm25.air.2018, "Date", "DAILY_AQI_VALUE", "Site.Name", "AQS_PARAMETER_DESC" , "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")


pm25.air.2019.subset <- select(pm25.air.2019, "Date", "DAILY_AQI_VALUE", "Site.Name", "AQS_PARAMETER_DESC" , "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")


#5 Mutate command is used to do any changes within the cells of a specific column.

summary(pm25.air.2018$AQS_PARAMETER_DESC)

pm25.air.2018.subset <- mutate(pm25.air.2018.subset, AQS_PARAMETER_DESC = "PM2.5")
pm25.air.2019.subset <- mutate(pm25.air.2019.subset, AQS_PARAMETER_DESC = "PM2.5")

#6 to save all processed data with write.csv file.

write.csv(ozone.air.2018.subset, row.names = FALSE,
          file = "./Data/Processed/EPAair_O3_NC2018_Processed.csv")
write.csv(ozone.air.2019.subset, row.names = FALSE,
          file = "./Data/Processed/EPAair_O3_NC2019_Processed.csv")
write.csv(pm25.air.2018.subset, row.names = FALSE,
          file = "./Data/Processed/EPAair_PM25_NC2018_Processed.csv")
write.csv(pm25.air.2019.subset, row.names = FALSE,
          file = "./Data/Processed/EPAair_PM25_NC2019_Processed.csv")


```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include only sites that the four data frames have in common:  
    
  "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue",  
  "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", 
  "West Johnston Co.", "Garinger High School", "Castle Hayne", 
  "Pitt Agri. Center", "Bryson City", "Millbrook School" 
  
  (the function `intersect` can figure out common factor levels - but it will include sites with missing site information, which you don't want...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package) #Ask question to John about this?
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r questions 7 thru 11}
#7 to combine the datasets with rbind function

air.ozone.pm25.2018_19 <- rbind(ozone.air.2018.subset, ozone.air.2019.subset, pm25.air.2018.subset, pm25.air.2019.subset)
dim(air.ozone.pm25.2018_19)

#8 Pipe command is used to avoid running function separately, we can run everything together through this command. 

air.ozone.pm25.2018_19_tidy <- 
  air.ozone.pm25.2018_19 %>% 
  filter(Site.Name == "Linville Falls" | Site.Name == "Durham Armory" | Site.Name == "Leggett" | Site.Name == "Hattie Avenue" | Site.Name == 
  "Clemmons Middle" | Site.Name =="Mendenhall School" | Site.Name =="Frying Pan Mountain" | 
  Site.Name =="West Johnston Co." | Site.Name == "Garinger High School" | Site.Name == "Castle Hayne" | 
  Site.Name == "Pitt Agri. Center" | Site.Name == "Bryson City" | Site.Name == "Millbrook School") %>% 
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>% 
  summarize(mean_AQI = mean(DAILY_AQI_VALUE),
            mean_LAT = mean(SITE_LATITUDE),
            mean_LON = mean(SITE_LONGITUDE)) %>%  
  mutate(Month = month(Date), Year = year(Date))
 
  
dim(air.ozone.pm25.2018_19_tidy) # 14,752 x 9

  
#9 Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

air.ozone.pm25.2018_19_tidy.spread <- air.ozone.pm25.2018_19_tidy %>% 
  pivot_wider(
    names_from = "AQS_PARAMETER_DESC", 
    values_from = "mean_AQI"
)

#10
dim(air.ozone.pm25.2018_19_tidy.spread) # dimentions are 8976 into 9

#11
write.csv(air.ozone.pm25.2018_19_tidy.spread, row.names = FALSE,
          file = "./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv") 

```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r questions 12 and 13}
#12
air.ozone.pm25.2018_19.summary <- air.ozone.pm25.2018_19_tidy.spread %>% 
 group_by(Site.Name, Month, Year) %>% 
  summarize(meanOzone = mean(Ozone),
            meanPM2.5 = mean(PM2.5)) %>% 
  drop_na(meanOzone)
#13
dim(air.ozone.pm25.2018_19.summary)
```

14. Why did we use the function `drop_na` rather than `na.omit`? Hint: replace `drop_na` with `na.omit` in part 12 and observe what happens with the dimensions of the summary date frame.

> Answer: na.omit function removes all the missing values from the data set, but with drop_na(), it removes NA from the mentioned variable, not from all. 

